---
title: "Reproducible Research - Peer Assessment 2"
author: "Tzachi Ezra Torf-Fulton"
date: "July 25, 2015"
output: html_document
---

# Impact of Severe Weather Events on Public Health and Economy in the United States

## Synopsis

In this report we analyze the impact of several weather events on the population
health and the economy, based on a public database maintained by the U.S. National 
Oceanic and Atmospheric Administration’s (NOAA) and between the years 1950 and 
2011. We will examine which events are the most harmful in terms of physical and 
financial damage.

Given the data we were able to find out that excessive heat is the main cause
for fatalities, followed by tornado. The later though, is by far the main cause 
for injuries. When we look at property damage we find that flood is the main
cause, where drought is the top reason for crop damage.

## Data Processing

### Load necessary libraries

```{r warning=FALSE, message=FALSE}
library(R.utils)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(stringr)
library(gridExtra)
```

### Download and extract the source file
 
```{r warning=FALSE, message=FALSE}
# check if raw csv file exist, if not download and upzip it
if (!file.exists("stormData.csv")) {
    if (!file.exists("stormData.csv.bz2")) {
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                      "stormData.csv.bz2", method = "curl")
    }
    # unzip file
    bunzip2("stormData.csv.bz2", remove = FALSE)
}
```

### Loading and preprocessing the data

We will use data frame table as our data holder as they are very powerful in 
processing big data sources. We will override the BGN_YEAR field with the year
value only, and will add to it the following fields (as the other are not) 
relevant for this analysis: BGN_YEAR, EVTYPE, FATALITIES, INJURIES, PROPDMG, 
PROPDMGEXP, CROPDMG, CROPDMGEXP

```{r cache=TRUE}
csvStormData <- tbl_df(read.csv("stormData.csv"))

# extract the year value and select the relevant coulmns for the analysis
csvStormData <- csvStormData %>%
    mutate(BGN_YEAR = year(mdy_hms(BGN_DATE))) %>%
    select(BGN_YEAR, EVTYPE, FATALITIES, INJURIES, 
                       PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

### Preview the number of events per year

To get a sense of the data, here is the total number of event records per
year.

```{r}
# group events by year
eventsPerYear <- csvStormData %>%
    group_by(BGN_YEAR) %>%
    summarise(count = n())

# plot bar graph
qplot(x=BGN_YEAR, y=count, data=eventsPerYear,  geom="bar", stat="identity", 
      fill=I("blue"), col=I("darkgreen"), alpha=I(.2), xlab="Year", 
      ylab="Number of Events", main="Total Number of Severe Weather Events by Year")
```

It is possible to see that until 1995 the number of events was very low, 
most likely due to a lack of good records. Therefore, in order to minimize the 
skew, in this report we will examine only events since 1995 (including).

```{r}
relevantData <- csvStormData %>%
    filter(BGN_YEAR >= 1995)
```

### Standardize PROPDMGEXP and CROPDMGEXP and calculate PROPDMG and CROPDMG

The next step in processing the data will be to normalize the raw values in
PROPDMGEXP and CROPDMGEXP to numeric fields. In case the raw field has numeric 
value we will keep it as is. Then we will convert H, K, M, B values to their
numeric exponent values 2, 3, 6, 9 (respectively). Empty values we be treated as 
0, and all the other characters will be considered as NA. The next step will be
to adjust PROPDMG and CROPDMG according to their raw value and their normalized 
exponent value.

```{r warning=FALSE}
# helper function to convert PROPDMGEXP and CROPDMGEXP raw values to numeric values
convertExp <- function (x) {
    ifelse(tolower(x)=="h", 2,
           ifelse(tolower(x)=="k", 3,
                  ifelse(tolower(x)=="m", 6, 
                         ifelse(tolower(x)=="b", 9, 
                                ifelse(str_trim(x)=="", 0, extract_numeric(x))))))
}

# filter by 1995 and convert exponent charaters to numbers
relevantData <- relevantData %>%
    mutate(PROPDMGEXP = convertExp(PROPDMGEXP)) %>%
    mutate(CROPDMGEXP = convertExp(CROPDMGEXP)) %>%
    mutate(PROPDMG = ifelse(is.na(PROPDMGEXP), NA, PROPDMG * (10 ^ PROPDMGEXP))) %>%
    mutate(CROPDMG = ifelse(is.na(CROPDMGEXP), NA, CROPDMG * (10 ^ CROPDMGEXP)))
```

### Breaking down the data

Now we can start looking at the health and economy impacts. For the health part,
we will split the data between fatalities and injuries and count the total number
of reported incidents by event type. For the economy part, we will split the data
between property ad crop, and count the total damage cost for each one. In this
report we will show the top 10 of each sub category.

```{r message=FALSE}
# select only health related data
healthData <- relevantData %>%
    select(BGN_YEAR, EVTYPE, FATALITIES, INJURIES)

# select only economy related data
economyData <- relevantData %>%
    select(BGN_YEAR, EVTYPE, PROPDMG, CROPDMG)

# extract fatalities from health related data, count total number of incidents
# based on event-type, arrange in decreasing order and select top 10
fatalities <- healthData %>%
    group_by(EVTYPE) %>%
    summarise(FATALITIES = sum(FATALITIES)) %>%
    select(EVTYPE, FATALITIES) %>%
    arrange(desc(FATALITIES)) %>%
    top_n(10)

# extract injuries from health related data, count total number of incidents
# based on event-type, arrange in decreasing order and select top 10
injuries <- healthData %>%
    group_by(EVTYPE) %>%
    summarise(INJURIES = sum(INJURIES)) %>%
    select(EVTYPE, INJURIES) %>%
    arrange(desc(INJURIES)) %>%
    top_n(10)
    
# extract property damage from economy related data, remove NAs, count total number of 
# incidents based on event-type, arrange in decreasing order and select top 10
propertyDamage <- economyData %>%
    select(BGN_YEAR, EVTYPE, PROPDMG) %>%
    na.omit() %>%
    group_by(EVTYPE) %>%
    summarise(PROPDMG = sum(PROPDMG)) %>%
    arrange(desc(PROPDMG)) %>%
    top_n(10)

# extract crop damage from economy related data, remove NAs, count total number of 
# incidents based on event-type, arrange in decreasing order and select top 10
cropDamage <- economyData %>%
    select(BGN_YEAR, EVTYPE, CROPDMG) %>%
    na.omit() %>%
    group_by(EVTYPE) %>%
    summarise(CROPDMG = sum(CROPDMG)) %>%
    arrange(desc(CROPDMG)) %>%
    top_n(10)
```

## Results

In the following tables we will summaries the top 10 causes for fatalities and 
injuries, as well as the top 10 causes for property and crop damage. 

### Fatalities and Injuries

```{r, message=FALSE}
# set event type as factor to maintain the order in the graph
fatalities$EVTYPE <- factor(fatalities$EVTYPE, levels=fatalities$EVTYPE)
#prepare the plot for fatalities
fatalitiesPlot <- qplot(x=EVTYPE, y=FATALITIES, data = fatalities, geom = "bar", 
                      stat="identity", fill=I("red"), col=I("black"), alpha=I(.4)) + 
    theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1, size=8)) + 
    scale_y_continuous("Number of Fatalities") + xlab("Event Type")

# set event type as factor to maintain the order in the graph
injuries$EVTYPE <- factor(injuries$EVTYPE, levels=injuries$EVTYPE)
#prepare the plot for injuries
injuriesPlot <- qplot(x=EVTYPE, y=INJURIES, data = injuries, geom = "bar", 
                      stat="identity", fill=I("orange"), col=I("black"), alpha=I(.4)) + 
    theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1, size=8)) + 
    scale_y_continuous("Number of Injuries") + xlab("Event Type")

# print the plots side by side
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2,
    main = "Total Number of Fatalities and Injuries by Severe Weather Events 
    Between 1995 and 2011 in ths US")
```

It is possible to see that Excessive Heat and Tornados are the main cause of
fatalities and injuries (respectively).

### Property and Crop Damage
```{r, message=FALSE}
# set event type as factor to maintain the order in the graph
propertyDamage$EVTYPE <- factor(propertyDamage$EVTYPE, levels=propertyDamage$EVTYPE)
#prepare the plot for property
propertyPlot <- qplot(x=EVTYPE, y=PROPDMG, data = propertyDamage, geom = "bar", 
                      stat="identity", fill=I("green"), col=I("black"), alpha=I(.4)) + 
    theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1, size=8)) + 
    scale_y_continuous("Property Damage in US Dollars") + xlab("Event Type")

# set event type as factor to maintain the order in the graph
cropDamage$EVTYPE <- factor(cropDamage$EVTYPE, levels=cropDamage$EVTYPE)
#prepare the plot for crop
cropPlot <- qplot(x=EVTYPE, y=CROPDMG, data = cropDamage, geom = "bar", 
                  stat="identity", fill=I("blue"), col=I("black"), alpha=I(.4)) + 
    theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1, size=8)) + 
    scale_y_continuous("Crop Damage in US Dollars") + xlab("Event Type")

# print the plots side by side
grid.arrange(propertyPlot, cropPlot, ncol=2,
    main = "Total Amount of Property and Crop Damaage in US Dollars by Severe 
    Weather Events Between 1995 and 2011 in ths US")
```

It is possible to see that Flood and Drought are the main cause of property and 
crop damage in US dollars (respectively).

## Conclusion Notes

While making the report, and as possible to see in the graphs above, the NOAA
graph is lacking standardization. Although we were able to "fix" the PROPDMGEXP
and CROPDMGEXP columns for this report, one can see that a significant work should
be done regarding the EVTYPE column. One of many examples can be seen at the
Crop Damage table, where Hurricane and Hurricane/Typhon are stated as different
events. This kind of phenomena causes the results to be skewed from the real
aggregated results per an event type.