---
title: "Reproducible Research - Peer Assessment 2"
author: "Tzachi Ezra Torf-Fulton"
date: "July 25, 2015"
output: html_document
---

# Impact of severe weather events on public health and economy in the United States

## Synopsis

In this report we analyse the impact of several weather events on the population
health and the economy, based on a public databased maintained by the U.S. National 
Oceanic and Atmospheric Administration’s (NOAA) and between the years 1950 and 
2011.

## Data Processing

##### Download the data

```{r warning=FALSE, message=FALSE}
library(R.utils)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(stringr)

# check if raw csv file exist, if not download and upzip it
if (!file.exists("stormData.csv")) {
    if (!file.exists("stormData.csv.bz2")) {
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                      "stormData.csv.bz2", method = "curl")
    }
    # unzip file
    bunzip2("stormData.csv.bz2", remove = FALSE)
}
```

##### Load the data

```{r cache=TRUE}
csvStormData <- tbl_df(read.csv("stormData.csv"))
```

```{r cache=TRUE}
csvStormData <- csvStormData %>%
    mutate(BGN_YEAR = year(mdy_hms(BGN_DATE))) %>%
    select(BGN_YEAR, EVTYPE, FATALITIES, INJURIES, 
                       PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

## Results

```{r}
eventsPerYear <- csvStormData %>%
    group_by(BGN_YEAR) %>%
    summarise(count = n())
qplot(x=BGN_YEAR, y=count, data=eventsPerYear,  geom="bar", stat="identity", 
      fill=I("blue"), col=I("darkgreen"), alpha=I(.2))
```

```{r warning=FALSE}
# helper function to convert PROPDMGEXP and CROPDMGEXP to numeric values
convertExp <- function (x) {
    ifelse(tolower(x)=="h", 2,
           ifelse(tolower(x)=="k", 3,
                  ifelse(tolower(x)=="m", 6, 
                         ifelse(tolower(x)=="b", 9, 
                                ifelse(str_trim(x)=="", 0, extract_numeric(x))))))
}

# filter by 1995 and convert exponent charaters to numbers
relData <- csvStormData %>%
    filter(BGN_YEAR >= 1995) %>%
    mutate(PROPDMGEXP = convertExp(PROPDMGEXP)) %>%
    mutate(CROPDMGEXP = convertExp(CROPDMGEXP)) %>%
    mutate(PROPDMG = ifelse(is.na(PROPDMGEXP), NA, PROPDMG * (10 ^ PROPDMGEXP))) %>%
    mutate(CROPDMG = ifelse(is.na(CROPDMGEXP), NA, CROPDMG * (10 ^ CROPDMGEXP)))
```

```{r}
healthData <- relData %>%
    select(BGN_YEAR, EVTYPE, FATALITIES, INJURIES)

economyData <- relData %>%
    select(BGN_YEAR, EVTYPE, PROPDMG, CROPDMG)

healthData
economyData
```