---
title: "Relationship Between Transmission Type and MPG"
author: "Tzachi Ezra Torf-Fulton"
date: "August 18, 2015"
output: pdf_document
---

# Relationship Between Transmission Type and MPG

## Executive Summary

In this report we analyzed the relationship between a set of variables 
and miles per gallon (MPG), based on a data that was extracted from the 1974 Motor 
Trend US magazine (mtcars - https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html). 
We focused in particular on quantifying the MPG difference between automatic and 
manual transmissions and determine whether an automatic or manual transmission 
better for MPG.

Given the data we were able to find out that 
excessive heat is the main cause
for fatalities, followed by tornado. The later though, is by far the main cause 
for injuries. When we look at property damage we find that flood is the main
cause, where drought is the top reason for crop damage.

## Data Processing
### Loading libraries and preprocessing the data
```{r warning=FALSE, message=FALSE}
library(ggplot2)
data(mtcars)
fmtcars<-mtcars
fmtcars$am <- factor(mtcars$am, labels=c("Automatic", "Manual"))
```

To get a sense of the relation between Transmission Type and MPG, we will start with
a general boxplot of these two variables.

```{r fig.height=2, fig.width=6, fig.align='center'}
p <- ggplot(fmtcars, aes(factor(am), x=am, y=mpg, fill=am))
p + geom_boxplot() + labs(x="Transmission Type", y="MPG") + 
    scale_fill_discrete(name="Transmission Type") + 
    theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1, size=8),
          axis.text.y = element_text(size=8), axis.title=element_text(size=9,face="bold")) 
```

Looking at this plot alone might lead to conclusion that manual transmission
has better mpg compare to automatic transmission. We will try to build a model
based on this assumption, and see how well it performs and whether transmission
type itself can be a good predictor on the mpg of a car.

### Single Variant Regression
In this first model we will try to fit the transmission type as a predictor for
a car's mpg.
```{r}
fit <- lm(mpg~am, data = fmtcars)
summary(fit)
```

From this model we can say that in average, given that all other factors do not
change, an automatic transmission underperform a manual one by 7.245 mpg. 
But looking carfuly we can see that the adjusted R-squared value is 33.85%, 
meaning that only 33.85% of the variation of the data can be explained by looking 
on transmission type alone.

This low adjusted R-squared value tells us that a model based on only on 
transmission type is not suffice to predict the mpg of a car.

```{r echo=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

### Multi Variant Regression

In order to decide which predictors we should pick for our model, let's take a
look at the correlation of mpg to each of the other variabls in the mtcars data set:
```{r}
sort(cor(mtcars)[1,])
```

From these results, it looks like there is a strong relationship between the 
weight of a car to its mpg. Also, from the graphs in Appendix-1, it is possible
to see that for all the variables, besides qsec, there is a strong separation
in values between manual and automatic transmission. This makes qsec (1/4 mile time) 
a very interesting variable as it is seems non-dependant in the other variables 
(in constrast for example to cyl and wt which seem very correlated).

We will add wt and qsec to our base model and see how it impacts the base model:

```{r}
fit1<-lm(mpg~am,data=fmtcars)
fit3<-update(fit1,mpg~am+wt)
fit5<-update(fit1,mpg~am+wt+qsec)
anova(fit1,fit3,fit5)
```

From the anova output, one can see that each of the extra variables is very significant 
(high RSS change, low P-value) to the model. It also means that we reject the
null hypothesis that our new modal is similar to the single-variable model.
Now that we belive we have a strong model we will check how well it performs.

```{r}
multiFit <- lm(mpg~am+wt+qsec, data = fmtcars)
summary(multiFit)
```

Our new model has an adjusted R-squared value of 83.36% which is significantly higher 
than our basic model. It also has a siginificant lower p-value, 1.21e-11, which means 
that extreme results using this model are far less common. Also, our multi-variant model 
residual plot (Appendix-2) shows that the points randomly dispersed around the horizontal axis, 
meaning a linear regression model is appropriate for the data. All these makes
this multi-variant model a stronger thesis.


### Regression Results Summary

As we take a look at the impact of transmission type on mpg using our last model,
we can see that manual transmission type makes an average car mpg higher by only 
2.9358 mpg. This is about 60% less our base analysis.

## Appendix

### Appendix-1
```{r fig.height=4}
variables <- c("cyl","disp","hp","drat","wt","qsec","vs","gear","carb")
plots <- list()
for (i in 1:length(variables)) {
    p<-ggplot(fmtcars, aes_string((variables[i]), "mpg", col="am")) + 
        geom_point() + theme(legend.position='none')
    plots[[i]]<-p
}
multiplot(plotlist = plots,  cols = 3)
```

The code for the multiplot function is taken from: 
http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/

### Appendix-2
```{r fig.height=4}
par(mfrow = c(2,2))
plot(multiFit)
```